Pythran: Enabling Static Optimization of Scientific Python Programs

### 2014-01-29 doc/papers/iop2014/paper.tex
Comment experiments & benchmarks.

The 4 hyper-threaded cores give an extra performance boost. Unfortunately, the
load is not sufficient to get more than an average 4x speed up compared to the
vectorized version.
% You complain about 4x because you only have 3.98x ???
In the end, Pythran generates a native module that performs
roughly 11\times faster than the original version.

@@

\begin{lstlisting}
def pi_estimate(darts):
    hits = 0
    "#omp parallel for reduction(+:hits)"
    for i in xrange(darts):
        x,y = random(), random()
        dist = sqrt(pow(x, 2) + pow(y, 2))
        if dist <= 1.0:
            hits += 1.0
    pi = 4 * (hits / DARTS)
    return pi
\end{lstlisting}
% why #? I thought string was enough?

@@

All benchmarks presented in this section are run on an hyper-threaded quad-core
i7, % FIXME sandy-brige, nehalem, what?
using examples shipped along with Pythran sources, available at
https://github.com/serge-sans-paille/pythran in the \texttt{pythran/test/cases}
directory. The Pythran version used is the \texttt{HEAD} of the \texttt{scipy2013} branch,
ShedSkin 0.9.2, PyPy 2.0 compiled with the \texttt{-jit} flag, CPython 2.7.3, Cython
0.19.1 and Numexpr 2.0.1. All timings are made using the \texttt{timeit} module,
taking the best of all runs. All C++ codes are compiled with g++ 4.8.2, using
the tool default compiler option, generally \texttt{-O2} plus a few optimizing flags
depending on the target.

Cython is not considered in most benchmarks, because to get an efficient
binary, one needs to rewrite the original code, while all the considered tools
are running the very same Python code that remains compatible with CPython. The
experiment was performed only to obtain a comparison with Numexpr.

Pystone is a Python translation of Whetstone, a famous floating point number
benchmarks that dates back to Algol60 and the 70's. Although non representative
of real applications, it illustrates the general performance of floating point
number manipulations. Table~\ref{tbl:pystone}  illustrates the benchmark
result for CPython, PyPy, ShedSkin and Pythran, using an input value of
\texttt{10**3}. Note that the original version has been updated to replace the user
class by a function call.

@@

    \begin{tabular}{|l|c|c|c|c|}
        \hline
     Tool    &  CPython    &   Pythran     &     PyPy   &  ShedSkin  \\
    \hline
     Timing  &  861ms      &   11.8ms      &     29.1ms &  24.7ms    \\
    \hline
     Speedup &  $\times$1         &   $\times$72.9       &    $\times$29.6   &  $\times$34.8     \\
    \hline
\end{tabular}

    \caption{Benchmarking result on the Pystone program.}
    \label{tbl:pystone}

@@

It comes at no surprise that all tools get more than decent results on this
benchmark. PyPy generates a code almost as fast % FIXME (dunno if it's efficient) 
as ShedSkin. Altough both
generate C++, Pythran outperforms ShedSkin thanks to a higher level generated
code. For instance all arrays are represented in ShedSkin by pointers to arrays
that likely disturbs the g++ optimizer, while Pythran uses a vector class wrapping
shared pointers.

Nqueen is a benchmark extracted from the former Unladen Swallow\footnote{\url{http://code.google.com/p/unladen-swallow/}} project. It
is particularly interesting as it makes an intensive use of non-trivial
generator expressions and integer sets. Table~\ref{tbl:nqueen} illustrates
the benchmark results for CPython, PyPy, ShedSkin and Pythran. The code had to
be slightly updated to run with ShedSkin because type inference in ShedSkin does
not support mixed scalar and \textbf{None} variables. The input value is \texttt{9}.

@@

    \begin{tabular}{|l|c|c|c|c|}
        \hline
     Tool    &  CPython    &   Pythran     &     PyPy   &  ShedSkin \\
    \hline
     Timing  &  1904.6ms   &   358.3ms     &    546.1ms &  701.5ms  \\
    \hline
     Speedup &  $\times$1         &    $\times$5.31      &    $\times$3.49   &  $\times$2.71    \\
    \hline
\end{tabular}
\caption{Benchmarking result on the NQueen program.}
\label{tbl:nqueen}

@@

It seems that compilers have difficulties to take advantage of high level
constructs such as generator expressions, as the overall speedup is not
breathtaking. Pythran benefits from the conversion to \texttt{itertools.imap} here,
while ShedSkin and PyPy rely on more costly constructs. A deeper look at the
Pythran profiling trace shows that more than half of the execution time is
spent allocating and deallocating a \texttt{set} used in the internal loop. There is a
memory allocation invariant that could be taken advantage of there, but none of
the compiler does so.

Hyantes\footnote{\url{http://hyantes.gforge.inria.fr}} is a geomatic application that exhibits typical usage of arrays
using loops instead of generalized expressions. It is helpful to measure the
performance of direct array indexing.

Table~\ref{tbl:hyantes} illustrates the benchmark result for CPython, PyPy,
ShedSkin and Pythran, when using lists as the data container. The output window
used is \texttt{100x100}.

@@

    \begin{tabular}{|l|c|c|c|c|}

        \hline
     Tool    &  CPython    &   Pythran     &     PyPy   &  ShedSkin  \\
    \hline
     Timing  &  1295.4ms   &   270.5ms     &    277.5ms &  281.5ms   \\
    \hline
     Speedup &  x1         &    x4.79      &    x4.67   &  x4.60     \\
    \hline
\end{tabular}
\caption{Benchmarking result on the hyantes kernel, list version.}
\label{tbl:hyantes}

@@

The speed ups are not amazing % FIXME (not formal enough, same for previous usage) 
for a numerical application. There are two
reasons for these poor speedups. First, the \texttt{hyantes} benchmark makes heavy
usage of trigonometric functions, and there is not much to gain there. Second, and
most important, the benchmark produces a big 2D array stored as a list of list,
hence the application suffers from the heavy overhead of converting them from C++
to Python. Running the same benchmark using Numpy arrays as core containers
confirms this assumption, as illustrated by Table~\ref{tbl:np-hyantes}. This
table also demonstrates the benefits of manual parallelization using OpenMP.
% Showing parallelization AND np.array at the same time confuse reader as we
% don't know how much each optimisation contribute in the speed up

@@

    \begin{tabular}{|l|c|c|c|c|}

        \hline
     Tool    &  CPython    &   Pythran     & Pythran+OpenMP   \\
    \hline
     Timing  &  450.0ms    &   4.8ms       &      2.3ms       \\
    \hline
     Speedup &  x1         &    x93.8      &    x195.7        \\
    \hline
\end{tabular}
\caption{Benchmarking result on the hyantes kernel, numpy version.}
\label{tbl:np-hyantes}
\end{table}

Finally, \texttt{arc\_distance} \footnote{The \texttt{arc\_distance} testbed is taken from to \url{https://bitbucket.org/FedericoV/numpy-tip-complex-modeling}.} presents a classical usage of Numpy expression. It
is typically more efficient than its loop alternative as all the iterations are
done directly in C. Its code is reproduced below:

\begin{lstlisting}
def arc_distance(theta_1, phi_1, theta_2, phi_2):
    """
    Calculates the pairwise arc distance
    between all points in vector a and b.
    """
    temp = (np.sin((theta_2 - theta_1) / 2) ** 2
        + np.cos(theta_1) * np.cos(theta_2)
          * np.sin((phi_2 - phi_1) / 2) ** 2)
    distance_matrix = 2 * np.arctan2(
            sqrt(temp), sqrt(1 - temp))
    return distance_matrix
\end{lstlisting}

Table~\ref{tbl:arc-distance} illustrates the benchmark result for CPython,
Cython, Numexpr and Pythran, using random input arrays of \texttt{10**6} elements.
Table~\ref{tbl:arc-distance-2} details the Pythran performance. Cython code
is written using the \texttt{parallel.prange} feature and compiled with
\texttt{-fopenmp -O2 -march=native}.

\begin{table}
    \centering
    \begin{tabular}{|l|c|c|c|c|}
    \hline
     Tool    &  CPython    &  Cython  &  Numexpr    & Pythran   \\
    \hline
     Timing  &  192.2ms    &  36.0ms  &    41.2ms   &  17.1ms   \\
    \hline
     Speedup &  x1         &  x5.33   &  x4.67      &  x11.23   \\
    \hline
\end{tabular}
% FIXME (You changed the order of the compilers, even though pythran is still first why?)
\caption{Benchmarking result on the arc distance kernel.}
\label{tbl:arc-distance}
\end{table}

\begin{table}
    \centering
    \begin{tabular}{|l|c|c|c|c|}
    \hline
    Tool & Pythran (raw) & Pythran (+AVX) & Pythran (+OMP) & Pythran (full)  \\
    \hline
    Timing & 186.3ms     &    75.4ms      &    41.1ms      &  17.1ms         \\
    \hline
    Speedup&    x1.03      &    x2.54       &    x4.67       &  x11.23         \\
    \hline
\end{tabular}
\caption{Detailed benchmarking result on the arc distance kernel for Pythran.}
\label{tbl:arc-distance-2}
\end{table}

It shows a small benefit from using expression templates on their own, most
% In fact there are small benefits for compiling this code using only pythran without AVX and openMP.
certainly because the loop control overhead is negligible compared to the
trigonometric functions. It gets a decent x2.5 speed-up when using AVX over not
using it. The benefit of OpenMP, although related to the number of cores, makes
an overall speedup greater than x11 compared to the original Numpy version, 
without
changing the input code. Almost on the opposite, Numexpr requires rewriting the 
input and does not achieve the same level of performance as Pythran when OpenMP
and AVX are combined.

Writing efficient Cython code requires more work than just typing the variable
declarations using Cython's specific syntax: it only takes advantage of
parallelism because we make it explicit. Without explicit parallelization,
the generated code runs around 176ms instead of 36ms. %FIXME speedup in x?
Cython does not generate
vectorized code, and \texttt{gcc} does not vectorize the inner loop, which 
explains the better result obtained with Pythran.

@@

Although Pythran focuses on a subset of Python and its standard library, many
optimizations opportunities are still present. Using a Domain Specific
Language(DSL) approach, one could use rewriting rules to optimize several
Python idioms. For instance, \texttt{len(set(x))} could lead to an optimized
\texttt{count\_uniq} that would iterate only once on the input sequence.

>>>>

The 4 hyper-threaded cores give an extra performance boost. In the end, Pythran
generates a native module that performs roughly $\times11$ faster than the
original version.

@@

\begin{lstlisting}
def pi_estimate(darts):
    hits = 0
    #omp parallel for reduction(+:hits)
    for i in xrange(darts):
        x,y = random(), random()
        dist = sqrt(pow(x, 2) + pow(y, 2))
        if dist <= 1.0:
            hits += 1.0
    pi = 4 * (hits / DARTS)
    return pi
\end{lstlisting}
% why #? I thought string was enough?

@@

All benchmarks presented in this section are run on an hyper-threaded quad-core
i7, % FIXME sandy-brige, nehalem, what?
using examples shipped along with
Pythran sources, available at \url{https://github.com/serge-sans-paille/pythran} in
the \texttt{pythran/test/cases} directory. The Pythran version used is the
\texttt{HEAD} of the \texttt{iop2014} branch, PyPy 2.0.2 compiled with the
\texttt{-jit} flag, CPython 2.7.3 and Cython 0.20. All timings are made using
the \texttt{timeit} module, taking the best of all runs. All C++ codes are
compiled with g++ 4.8.2, using the tool default compiler option, generally
\texttt{-O3} plus a few optimizing flags depending on the target. To make a
fair comparison, no parallelism annotations are added to any code.

Provided each module has been properly compiled, each experiment is run using
the following command from a shell:

\begin{lstlisting}[language=sh]
python -m timeit -s "${SETUP}" "${RUN}"
\end{lstlisting}

\noindent where \texttt{\$SETUP} and \texttt{\$RUN} depend on the benchmark and
\texttt{pypy} is substituted to \texttt{python} when needed. The
\texttt{\$SETUP} and \texttt{\$RUN} values are available to the reader on the
Git repository in the \texttt{doc/papers/iop/xp/run\_benchamrks.sh} file with
all the experimental setup.

The same source is used for each benchmark, to the exception of Cython sources
which needed a manual conversion. For these sources, bound checking and wrap
around are always disable, all variables are declared using the proper C or C++
type and all mathematical functions in use come from \texttt{libm}. Some calls
to the \texttt{zeros\_like} function also have been replaced by calls to
\texttt{zeros} for the PyPy version.

\subsection{NQueens}

``Nqueens'' is a benchmark extracted from the former Unladen
Swallow\footnote{\url{http://code.google.com/p/unladen-swallow/}} project. It
is particularly interesting as it makes an intensive use of non-trivial
generator expressions and integer sets.

Table~\ref{tbl:nqueen} illustrates the benchmark results for CPython, Cython,
PyPy and Pythran.  It showcases the fact that Pythran benefits a lot from
high-level transformations, for instance \emph{lazy evaluation}, which prevents
the many allocation / deallocation triggered by Cython and PyPy runs.  A deeper
look at the Pythran profiling trace shows that more than half of the execution
time is spent allocating and deallocating a \texttt{set} used in the internal
loop. There is a memory allocation invariant that could be taken advantage of
there, but none of the compiler does so.

@@

    \begin{tabular}{|l|c|c|c|c|}
        \hline
     Tool    &  CPython    &   Cython     &     Pythran   &  PyPy \\
    \hline
    Timing  &  181ms   &   42.3ms     &    \textbf{22.5ms} &  51.3ms  \\
    \hline
    Speedup &  $\times$1         &    $\times$4.27      &    \textbf{$\times$8.04}   &  $\times$3.52    \\
    \hline
\end{tabular}
\caption{Benchmarking result on the NQueen program.}
\label{tbl:nqueen}

@@

\subsection{Rosen Der}

The ``rosen der'' kernel computes a complex numpy expression that accesses
data through a one dimensional slice. It is interesting to evaluate how much of
the expression can be compiled to regular, efficient loop while avoiding
temporaries.  Table~\ref{tbl:rosen} illustrate the benchmark results for
CPython, Cython, PyPy and Pythran. Cython slightly outperforms Pythran on that
test case, but Cython code had to be manually transformed into an explicit
loop. Moreover, activating automatic parallelization for Pythran yields an
additional $\times4$ speedup without a change to the input, while Cython needs
manual code annotation to achieve a similar result. It seems that PyPy support
for numpy expression is still note mature for numpy expression, but there is a
strong ungoing effort of the community to fix this.

@@

    \begin{tabular}{|l|c|c|c|c|}
        \hline
     Tool    &  CPython    &   Cython     &     Pythran   &  PyPy \\
    \hline
    Timing  &  22.3ms   &   \textbf{3.39ms}     &    3.81ms &  51.3ms  \\
    \hline
    Speedup &  $\times$1         &    \textbf{$\times$6.57}      &    $\times$5.85   &  $\times$0.43    \\
    \hline
\end{tabular}
\caption{Benchmarking result on the Rosen der program.}
\label{tbl:rosen}

@@

\subsection{Growcut}

The ``growcut'' application is an image processing code where all array indexing
are explicit. In that sense, this benchmark is complementary of the previous
one, as it evaluates how well explicit array access are compiled.
Table~\ref{tbl:growcut} summarizes the results. All compiler yield impressive
speedups on that benchmark, but this is is merely due to the fact that
accessing individual elements is extremely slow in numpy, and the direct C
translation of the Python code is very close to what a C developer would write.
Pythran appears to outperforms Cython once all access are explicit, most
certainly due to cleaner code generation.

@@

    \begin{tabular}{|l|c|c|c|c|}
        \hline
     Tool    &  CPython    &   Cython     &     Pythran   &  PyPy \\
    \hline
    Timing  &  398ms   &   991us     &    \textbf{526us} &  5.54ms  \\
    \hline
    Speedup &  $\times$1         &    $\times$401.6      &    $\times$756.7   &  $\times$71    \\
    \hline
\end{tabular}
\caption{Benchmarking result on the growcut program.}
\label{tbl:growcut}

@@

\subsection{Blacksholes}

The ``blacksholes'' application is a benchmarks that showcases the
translation cost from the Python space to native space. Indeed we forced the
use of a \texttt{list} as input of the kernel instead of a numpy array. As a
consequence there is a costly translation step from a structure that contains
an array of pointers to generic Python objects to a dense structure involved.
Results are shown in Table~\ref{tbl:black}. They show that conversion code is
less efficient in Cython than in Pythran. PyPy also outperforms Cython, because
it does not suffer from a similar conversion penality.

@@

    \begin{tabular}{|l|c|c|c|c|}
        \hline
     Tool    &  CPython    &   Cython     &     Pythran   &  PyPy \\
    \hline
    Timing  &  490us   &   74.7us     &    \textbf{34.8us} &  37.3us  \\
    \hline
    Speedup &  $\times$1         &    $\times$6.55      &    $\times$14.1   &  $\times$13.1    \\
    \hline
\end{tabular}
\caption{Benchmarking result on the blacksholes program.}
\label{tbl:black}

@@

Although Pythran focuses on a subset of Python and its standard library, many
optimizations opportunities are still available. Using a Domain Specific
Language (DSL) approach, one could use rewriting rules to optimize several
Python idioms. For instance, \texttt{len(set(x))} could lead to an optimized
\texttt{count\_uniq} that would iterate only once on the input sequence.
